Generating and visualizing predictor effects???

### 2021-09-22 variable_predictions.Rnw
Autocommit

\def\code#1{\texttt{#1}}

@@

Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function (in the case of simple linear model), where $\boldmu$ is the expected value of response variable $y$. Let $\hat{\bbeta}$ be the estimate of $\bbeta$, together with the estimated covariance matrix $V(\hat{\bbeta})$ of $\hat{\bbeta}$. Let $\mathbf{X^*}$ be the model matrix, inheriting most of its key properties, for example transformations on predictors and interactions from the model matrix, $\mathbf{X}$. Then the prediction $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the variable effect for the focal predictor in question.

@@

What about the confidence intervals (CI)? The limits of the confidence intervals are points, not mean values. In principle, every observation/set of non-focal predictors has a different CI.

>>>>

\def\code#1{\texttt{#1}}
\let\proglang=\textsf

@@

Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function (in the case of simple linear model), where $\boldmu$ is the expected value of response variable $y$. Let $\hat{\bbeta}$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\hat{\bbeta})$ of $\hat{\bbeta}$. Let $\mathbf{X^*}$ be the model matrix, inheriting most of its key properties, for example transformations on predictors and interactions from the model matrix, $\mathbf{X}$. Then the prediction $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the variable effect for the focal predictor in question.

@@

What about the confidence intervals (CI)? The limits of the confidence intervals are points, not mean values. In principle, every observation/value of focal predictors has a different CI. The traditional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \Sigma \bX^{\star\top})$ \citep{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution. This approach incorporates all the uncertainties -- including the uncertainty due to non-focal predictors.  But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$?

Currently, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from the \emph{non-focal} predictors when computing the CIs. A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of \emph{non-focal} terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix. This only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and even much complicated when the predictors are categorical. We first describe this variance-covariance based approach and then discuss our proposed method which is based on \emph{model centering} and does not require input predictors to be scaled prior to model fitting. Model center is found by averaging the variables over the model matrix, $\bX^\star$.

\subsection{Variance-covariance}

The computation of $\hat{\eta}^\star$ remains the same as described above. However, to compute $\sigma$, $\Sigma$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are assigned zero) variances of non-focal terms. Although this is the simplest approach, it requires centering of continuous predictors prior to model fitting and proper handling of categorical predictors. One way to center predictors is to create a design matrix, $\bX$, and then compute $\bX_{c} = \bX - \bar{\bX}$.

\subsection{Centered model matrix}

Consider centered model matrix $\bX^{\star}_{c} = \{\bX_f^\star, \nset{{\bX}^\star} - \bar{\nset{{\bX^\star}}}\}$. It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero. Consequently the uncertainty due to non-focal predictors are all zeroed-out in the computation of $\sigma$. More generally, centered design matrix, $\bX^{\star}_{c}$, impacts on the estimated value of the intercept and its associated variance but not the slopes. Thus since non-focal terms in $\bX^{\star}_{c}$ are all zero, it does not matter what their corresponding values are in the variance-covariance matrix. Hence, we can compute CIs from non-centered predictors (in other words, fitted models with predictors in their natural scales).

\subsection{Simulation examples}

In this section, we illustrate variable effects in the context of linear models involving continuous and categorical predictors with or without interactions. We compare the expected to the known simulation "truth". We compare predictions when non-focal predictors averaging and when population-based approaches are used. Later on, we illustrate and compare our approaches for describing uncertainty in the predictions to existing major \textbf{R} packages for prediction.

\subsection{Simple linear model}

Consider a simple simulation
\begin{align*}
y &= \beta_0 + \beta_1\mathrm{x_1} + \beta_2\mathrm{x_2} + \epsilon \\
\mathrm{x_1} &\sim \mathrm{Normal}(0.2, 1) \\
\mathrm{x_2} &\sim \mathrm{Normal}(0, 1) \\
\epsilon &\sim \mathrm{Normal}(0, 1) \\
\beta_0 &= 1.5 \\
\beta_{\mathrm{x_1}} &= 1.0 \\
\beta_{\mathrm{x_2}} &= 2
\end{align*}
for $10000$ observations.

```{r simple_linear_sim, echo=FALSE}
sim_linear_df <- simplesim(N=1e4, link_scale=TRUE)
colnames(sim_linear_df) <- c("x1", "x2", "y")
true_prop_linear <- mean(sim_linear_df$y)
```
